### 2016-03-25 paper.tex
Incorporated Ash's edits

% Abstract: 150 words or less, extra 0.5in left and right margin.
% Note that all text is justified.
%
\begin{adjustwidth}{0.5in}{0.5in}
  \textbf{Abstract:} We are inspired by the educational potential of
  the board game \textit{Tales of the Arabian Nights}, which is a
  popular game among enthusiasts that teaches the player about the
  eponymous folk tale and the culture around it. Consider this game
  and others like it, we identify the characteristics that define a
  genre of \textit{culture-narration games}, which we consider to have
  untapped potential for educational and transformative games.
  We describe a design experiment through which a multidisciplinary team
  followed an iterative and incremental process, in collaboration
  with a community partner, to investigate the potential of this genre.
  The result is a game that uses a theme of monsters from around the
  world to teach cultural empathy.
  This pilot project reveals both the promise and the several complications
  with the genre, which lead to recommendations for future work.
\end{adjustwidth}

@@

\textit{Tales of the Arabian Nights}~\citep{Goldberg2009} is a 
tabletop board game based on the eponymous folk tale,
originally published in 1985~\citep{Goldberg1985} and
released with many revisions in 2009.
It is popular among board game hobbyists, having a board game
rank of~239 (thematic rank~75) on Board Game Geek as of this writing,
and Shut Up \& Sit Down rank it as the ninth best game of all 
time~\citep{ShutUp2015}.
In this game, players control a character within a mythical Arabian
setting, exploring the known world to accumulate Story and
Destiny points. The winner is the one who is able to meet their 
Story and Destiny point goal while having a successful encounter
in Baghdad---the City of Peace.

@@

This property of \totan{} reflects several characteristics
of its source material---a book that ``changed the world on a scale
unrivalled by any other literary text''~\citep[p.1]{Makdisi2008}.
Like the original stories, \totan{} is a collection of shorter stories
that are sometimes linked together and sometimes not.
The characters in the folk tales are often victims of fate with
very little agency over their own encounters. The games rules follow
cultural norms expressed in the folk tales:
you cannot win while sex-changed or on pilgrimage, and both marriage
and children are great blessings---unless you have an ugly baby,
which is shameful.
Baghdad is the most important city in the world, where your adventure starts
and ends.
Furthermore, many stories in \totan{} come directly from the tales,
such as the Sindbad's escape from the valley of diamonds or Aladdin's
being trapped in a magic cave.

Playing \totan{}  inspired members of our team to read translations
of the original tales,
particularly  the youth-friendly translation by \citep{Neil1994},
which made us recognize the cleverness of the design
and its elegant dovetailing with the source material.
Indeed, although the game was not explicitly designed as a learning game,
we find in it exhibits many of the learning game design principles
established by \citet{Klopfer2009}.
Furthermore, in reading the texts, we recognized that we had already
learned elements of the stories and the culture through playing the game,
although not always consciously or explicitly.
This inspired the following analysis, in which we tease apart the
various elements of \totan{}, considering them from both game design
and learning design points of view, and then share the results of a pilot
project to create a technology-enhanced game within the same genre.

\section{Defining the genre}

@@

\item \textit{Reward players for decisions that reflects
    cultural understanding.}  The knowledge that a player brings to
  bear on decisions is not just knowledge of in-game systems. 
  A player can predict that choosing to fight a Powerful Ifrit 
 without the Weapon Use skill will not end well,
  but this prediction is not made by having memorized the 
  \textit{Book of Tales}.
  Rather, this prediction draws upon a cultural understanding of power,
  spirits, violence, and much more.
  Note that the game rewards player decisions without resorting to
  clear-cut ``right'' and ``wrong'' (``moral/immoral'',
  ``light side/dark side'', ``paragon/renegade'') choices.  There
  remain unexpected twists: perhaps fighting the Powerful Ifrit without
  combat skills makes him respect you and grant a boon.  \totan{},
  after all, is a game that allows the player to choose to ``drink'' a
  Violent Storm or ``enter'' a Mysterious Artifact.  While the world
  remains consistent, it also embraces the potential for surprising
  results.

@@

To establish these as characteristics of a genre and not just one game,
we consider another game that possesses them: 
\textit{Agents of SMERSH}~\citep{Maxwell2012}.
\smersh{} is a cooperative game set in the Cold War, with players as 
covert agents trying to stop the enigmatic Dr.\ Lobo. 
Players move across the world collecting resources and battling henchmen.
While the stories are inspired by spy movies and novels,
the game mechanics are clearly derived from \totan{}: flip a card to read an
encounter title, choose a reaction from a table, hear an introduction,
and then get different results based on skills.
Although \smersh{} uses dice for encounter resolution, the feedback is still
primarily narrative, as illustrated with sample encounters in
Figure~\ref{fig:encounters}.

For contrast, we also consider \textit{Above and Below}~\citep{Laukat2015}, a
worker-management game of city-building.  
%\textit{Above and Below} is primarily about building an
%economic engine that generates victory points faster than your
%opponents within a fixed number of rounds,
Underground exploration includes  player narration
from an encounter book as in \totan{} and \smersh{}, which at first
glance may appear to categorize it as a culture-narration game.
However, in \textit{Above and Below}, reaction choices are thin
wrappers around risk-and-reward balance:
an option is essentially a gamble on being able to roll better
results on the dice, with higher numbers always having better results.
The feedback is primarily in terms of formal rewards, with few
encounters providing any narrative feedback, as illustrated in
the representative encounter of Figure~\ref{fig:encounters}.
Hence, although \textit{Above and Below} includes an encounter book
that players read to each other, it does not satisfy the characteristics
to be in the culture-narration game genre.

@@

Cultural empathy is distinct from \textit{cultural
literacy}, which prepares individuals to participate in another culture
based on the information of the culture that they have been 
given~\citep[p.165]{Hirsch1983}. Being culturally literate is having the
ability to be able to participate in another culture that you are
``dropped into,'' similar to how being literate is having the ability to
understand or write a given work of text.  \totan{} focuses on a certain
level of literacy of the given culture. Players are rewarded for their
cultural understanding and their ability to ``perform'' in the given
culture because of the information they have---a form of what
\citet{Travis2011} calls \textit{practomime}.
This is demonstrated in
the example above, understanding the cultural views of marriage and
sex-change have an important role in the game. However, cultural
empathy makes understanding the main point, not
performance. \totan{} relies on an understanding of cultural values as
opposed to just focusing on some of their values consequential
norms. Therefore, though \totan{} expects some sense of cultural
literacy, it focuses more on cultural empathy. When a player walks
away from the game they would not have gained the ability to
participate in a Middle Eastern culture,
although they will have gained some understanding of that culture. 
%This
%is why our emphasis in creating the game has relied more on cultural
%empathy than cultural literacy. We focus more on a mindset of a
%culture, than about facts about the culture. 
Hence, from a learning game design perspective,
cultural empathy is a means to the end of cultural
literacy. However, outside of the game, focusing on cultural literacy
within a semi-fictional world is the means for cultivating
cultural empathy.

@@

In order to explore the potentials within this genre,
we formed a team to conduct a design experiment and pilot study.
The team adopted a constructionist perspective~\citep{Papert2001},
that our understanding of the genre would be improved by engaging
with design, development, and evaluation---building prototypes
as a way to understand them.
We formed a partnership with \textit{[omitted for blind review]}, 
a non-lending branch that 
provides community, training, and technology services in a
low socioeconomic area of our city.
The youth patrons of the library tend to be 4rd--6th grade 
``latchkey kids'' who 
see the library as an after-school destination to visit with friends, 
to participate in programs, and to access technology.

Conversations with program coordinators about culture-narration games
revealed that despite other differences, the youth patrons shared a
fascination with monsters, horror, and the macabre.  This answer in
some ways seemed unsatisfactory: horror and empathy---cultural or
otherwise---do not seem to be a good match. Indeed,
to enjoy horror requires an antithetical dehumanization of its subjects.
However, the love of horror and the monstrous can be
seen throughout our culture, in all forms of media. Indeed, monsters have
been a hook for narratives throughout history. They have attracted
people of various walks of life, by playing into shared human and cultural
fears.

@@

Our team therefore chose monsters as a key dramatic element for our 
design experiments. They were not only enticing to our target
audience---first observed by our community partner and later proven
through play testing---but Cohen's theory also 
suggests that they also provide an excellent passage for
encouraging empathy. Monsters represent
various cultures' sameness, in base forms of fear, yet difference, in
the ways that fear becomes specific and embodied. By encouraging an
understanding of both difference and sameness between cultures, they
encourage individuals to try to see something from someone else's
viewpoint.

@@

The title and theme of our game changed between iterations as we came
to better understanding of the interplay between genre, rules, and our
players.  The final version---\textit{Traveler's Story: Monster
  Tales}---positioned two players as rival authors seeking inspiration
from monsters around the world. The players travel across Africa,
Asia, Europe, Oceania, and North and South America to encounter
monsters from local folklore.  Encounters are patterned after \totan{}
as illustrated in Figure~\ref{fig:encounters}, and the game exhibits
the five characteristics identified for cultural-narration games.

>>>>

% Abstract: 150 words or less, extra 0.5in left and right margin.
% Note that all text is justified.
%
\begin{adjustwidth}{0.5in}{0.5in}
  \textbf{Abstract:} We are inspired by the educational potential of
  the board game \textit{Tales of the Arabian Nights}, a
  popular game based on the eponymous folk tale. Considering this game
  and others like it, we identify the characteristics that define a
  genre of \textit{culture-narration games}, which we consider to have
  untapped potential for educational and transformative games.
  We describe a design experiment through which a multidisciplinary team
  followed an iterative and incremental process, in collaboration
  with a community partner, to investigate the potential of this genre.
  The result is a game that uses a theme of monsters from around the
  world to teach cultural empathy.
  This pilot project reveals both the promise and several complications
  with the genre, which lead to recommendations for future work.
\end{adjustwidth}

@@

\textit{Tales of the Arabian Nights}~\citep{Goldberg2009} is a 
tabletop board game based on the eponymous folk tale. It was
originally published in 1985~\citep{Goldberg1985}, and it was
re-released with many revisions in 2009.
It is popular among board game hobbyists:
on Board Game Geek, it has a rank of~239 (thematic rank~75),
and Shut Up \& Sit Down rank it as the ninth best game of all 
time~\citep{ShutUp2015}.
In this game, players control a character within a mythical Arabian
setting, exploring the known world to accumulate Story and
Destiny points. The winner is the player who is able to meet their 
Story and Destiny point goal while having a successful encounter
in Baghdad---the City of Peace.

@@

This property of \totan{} reflects several characteristics
of its source material---a book that ``changed the world on a scale
unrivalled by any other literary text''~\citep[p.1]{Makdisi2008}.
Like the original stories, \totan{} is a collection of shorter stories
that are sometimes linked together and sometimes not.
The characters in the folk tales are often victims of fate with
very little agency over their own encounters. The games rules follow
cultural norms expressed in the folk tales:
you cannot win while sex-changed or on pilgrimage, and both marriage
and children are great blessings---unless you have an ugly baby,
which is shameful.
Baghdad is the most important city in the world, where your adventure starts
and ends.
Furthermore, many stories in \totan{} come directly from the tales,
such as the Sindbad's escape from the valley of diamonds or Aladdin's
entrapment in a magic cave.

Playing \totan{}  inspired members of our team to read translations
of the original tales,
particularly  the youth-friendly translation by \citep{Neil1994},
which revealed to us the cleverness of the design
and its elegant dovetailing with the source material.
Indeed, although the game was not explicitly designed as a learning game,
we find it exhibits many of the learning game design principles
established by \citet{Klopfer2009}.
Furthermore, in reading the texts, we recognized that we had already
learned elements of the stories and the culture through playing the game,
although not always consciously or explicitly.
This inspired the following analysis, in which we tease apart the
various elements of \totan{}, considering them from both game design
and learning design points of view, and then share the results of a pilot
project to create a technology-enhanced game within the same genre.

\section{Defining the Genre}

@@

\item \textit{Reward players for decisions that reflects
    cultural understanding.}  The knowledge that a player brings to
  bear on decisions is not just knowledge of in-game systems. 
  A player can predict that choosing to fight a Powerful Ifrit 
 without the Weapon Use skill will not end well,
  but this prediction is not made by having memorized the 
  \textit{Book of Tales}.
  Rather, this prediction draws upon a cultural understanding of power,
  spirits, violence, and much more.
  Note that the game rewards player decisions without resorting to
  clear-cut ``right'' and ``wrong'' (``moral/immoral'',
  ``light side/dark side'', ``paragon/renegade'') choices.  There
  remain unexpected twists: perhaps fighting the Powerful Ifrit without
  combat skills makes him respect you and grant a boon. 
  While the world
  remains consistent, it also embraces the potential for surprising
  results.

@@

To establish these as characteristics of a genre and not just one game,
we consider another game that possesses them: 
\textit{Agents of SMERSH}~\citep{Maxwell2012}.
\smersh{} is a cooperative game set in the Cold War, with players as 
covert agents trying to stop the entigmatic Dr.\ Lobo---a mad scientist
bent on world domination.
Players move across the world collecting resources and battling henchmen.
While the stories are inspired by spy movies and novels,
the game mechanics are clearly derived from \totan{}: flip a card to read an
encounter title, choose a reaction from a table, hear an introduction,
and then get different results based on skills.
Although \smersh{} uses dice for encounter resolution, the feedback is still
primarily narrative, as illustrated with sample encounters in
Figure~\ref{fig:encounters}.

For contrast, we also consider \textit{Above and Below}~\citep{Laukat2015}, a
worker-management game of city-building.  
%\textit{Above and Below} is primarily about building an
%economic engine that generates victory points faster than your
%opponents within a fixed number of rounds,
Underground exploration includes  player narration
from an encounter book as in \totan{} and \smersh{}, which at first
glance may appear to categorize it as a culture-narration game.
However, in \textit{Above and Below}, reaction choices are thin
wrappers around risk-and-reward balance:
an option is essentially a gamble on being able to roll better
results on the dice, with higher numbers always having better results.
The feedback is primarily in terms of formal rewards, with few
encounters providing any narrative feedback, as illustrated in
the representative encounter of Figure~\ref{fig:encounters}.
Hence, although \textit{Above and Below} includes an encounter book
that players read to each other, it does not satisfy the characteristics
of the culture-narration game genre.

@@

Cultural empathy is distinct from \textit{cultural
literacy}, which prepares individuals to participate in another culture
based on the information of the culture that they have been 
given~\citep[p.165]{Hirsch1983}. Being culturally literate means having the
ability to participate in another culture that you are
``dropped into,'' similar to how being literate is having the ability to
understand or write a given work of text.  \totan{} focuses on a certain
level of literacy of the given culture. Players are rewarded for their
cultural understanding and their ability to ``perform'' in the given
culture because of the information they have---a form of what
\citet{Travis2011} calls \textit{practomime}.
This is demonstrated in
the example above, understanding the cultural views of marriage and
sex-change have an important role in the game. However, cultural
empathy makes understanding the main point, not
performance. \totan{} relies on an understanding of cultural values as
opposed to just focusing on some of their values consequential
norms. Therefore, though \totan{} expects some sense of cultural
literacy, it focuses more on cultural empathy. When a player walks
away from the game they would not have gained the ability to
participate in a Middle Eastern culture,
although they will have gained some understanding of that culture. 
%This
%is why our emphasis in creating the game has relied more on cultural
%empathy than cultural literacy. We focus more on a mindset of a
%culture, than about facts about the culture. 
Hence, from a learning game design perspective,
cultural empathy is a means to the end of cultural
literacy. However, outside of the game, focusing on cultural literacy
within a semi-fictional world is the means for cultivating
cultural empathy.

@@

In order to explore the potentials within this genre,
we formed a team to conduct a design experiment and pilot study.
The team adopted a constructionist perspective~\citep{Papert2001},
that our understanding of the genre would be improved by engaging
with design, development, and evaluation---building prototypes
as a way to understand them.
We formed a partnership with \textit{[omitted for blind review]}, 
a non-lending branch that 
provides community, training, and technology services in a
low socioeconomic area of our city.
The youth patrons of the library tend to be 4rd--6th grade 
``latchkey kids'' who 
see the library as an after-school destination to visit with friends, 
participate in programs, and access technology.

Conversations with program coordinators about culture-narration games
revealed that despite other differences, the youth patrons shared a
fascination with monsters, horror, and the macabre.  This answer in
some ways seemed unsatisfactory: horror and empathy---cultural or
otherwise---do not seem to be a good match. Indeed,
to enjoy horror requires an antithetical dehumanization of its subjects.
However, the love of horror and the monstrous can be
seen throughout our culture, in all forms of media. Indeed, monsters have
been a hook for narratives throughout history.

@@

Our team therefore chose monsters as a key dramatic element for our 
design experiments. They were not only enticing to our target
audience---first observed by our community partner and later proven
through play testing---but Cohen's theory also 
suggests that they also provide an excellent passage for
developing empathy. Monsters represent
various cultures' sameness, in base forms of fear, yet difference, in
the ways that fear becomes specific and embodied. By engendering an
understanding of both difference and sameness between cultures, they
encourage individuals to try to see something from someone else's
viewpoint.

@@

The title and theme of our game changed between iterations as we came
to a better understanding of the interplay between genre, rules, and our
players.  The final version---\textit{Traveler's Story: Monster
  Tales}---positioned two players as rival authors seeking inspiration
from monsters around the world. The players travel across Africa,
Asia, Europe, Oceania, and North and South America to encounter
monsters from local folklore.  Encounters are patterned after \totan{}
as illustrated in Figure~\ref{fig:encounters}, and the game exhibits
the five characteristics identified for cultural-narration games.
